\chapter{Abstract}
Kubernetes does not impose a specific network plugin, forcing teams to select a \emph{Container Network Interface} (CNI) without objective comparative metrics. This situation leads to infrastructure oversizing (CPU/RAM) and incomplete security postures. This thesis builds reproducible empirical evidence —infrastructure as code, automated benchmarks, and a multi-criteria decision analysis (MCDA) model— to guide this selection. Flannel, Calico, Cilium, and Antrea are evaluated on a high-availability K3s cluster on DigitalOcean, measuring TCP throughput, establishment latency, retransmissions, and CPU/RAM overhead of the CNI agent, as well as the impact of Zero Trust network policies. The results feed a parameterizable MCDA model by use profile (Fintech/security, Streaming/performance, and IoT/limited resources) and a web prototype that displays the recommendations.
The main findings show that Calico achieved the highest TCP throughput (1804.40 Mbps) and the lowest connection-establishment latency; that Flannel —the default CNI in K3s— accepts NetworkPolicies through the Kubernetes API but does not enforce them in its data plane, constituting a security false positive; and that Cilium was the only plugin capable of enforcing layer-7 (HTTP) policies. The model recommends Cilium for the Fintech profile, and Flannel for the Streaming and IoT profiles due to its stability and smaller resource footprint.

\vspace{1cm}
\textbf{Keywords:} Kubernetes, CNI, NetworkPolicy, Zero Trust, MCDA, QoS, eBPF.
\newpage