\begin{thebibliography}{1}

\bibitem[1]{ref1}
``EndpointSlices,'' Kubernetes. [En línea]. Disponible en: \url{https://kubernetes.io/docs/concepts/services-networking/endpoint-slices/}. [Consultado: 21-ago-2025].

\bibitem[2]{ref2}
``Cluster networking,'' Kubernetes. [En línea]. Disponible en: \url{https://kubernetes.io/docs/concepts/cluster-administration/networking/}. [Consultado: 21-ago-2025].

\bibitem[3]{ref3}
``Virtual IPs and Service proxies,'' Kubernetes. [En línea]. Disponible en: \url{https://kubernetes.io/docs/reference/networking/virtual-ips/}. [Consultado: 21-ago-2025].

\bibitem[4]{ref4}
``Network plugins,'' Kubernetes. [En línea]. Disponible en: \url{https://kubernetes.io/docs/concepts/extend-kubernetes/compute-storage-net/network-plugins/}. [Consultado: 21-ago-2025].

\bibitem[5]{ref5}
D. E. Eisenbud \emph{et al.}, ``Maglev: A fast and reliable software network load balancer,'' \emph{USENIX NSDI}, 2016.

\bibitem[6]{ref6}
F. Gomes, P. Rego, y F. Trinta, ``A systematic mapping study on observability of microservices-based applications: fundamentals, classifications, and challenges,'' \emph{Computing}, vol. 107, núm. 9, 2025.

\bibitem[7]{ref7}
Z. Kang, K. An, A. Gokhale, y P. Pazandak, ``A comprehensive performance evaluation of different Kubernetes CNI plugins for edge-based and containerized publish/subscribe applications,'' Vanderbilt University, 2021.

\bibitem[8]{ref8}
G. Koukis, S. Skaperas, I. A. Kapetanidou, L. Mamatas, y V. Tsaoussidis, ``Performance evaluation of Kubernetes networking approaches across constraint edge environments,'' \emph{arXiv} [cs.NI], 2024.

\bibitem[9]{ref9}
MDPI. [En línea]. Disponible en: \url{https://www.mdpi.com/2079-9292/13/19/3972}. [Consultado: 21-ago-2025].

\bibitem[10]{ref10}
``Elevating Kubernetes network security through Cilium deployment,'' FH Joanneum. [En línea]. Disponible en: \url{https://epub.fh-joanneum.at/obvfhjhs/download/pdf/11499653}. [Consultado: 21-ago-2025].

\bibitem[11]{ref11}
W. Tu, Y.-H. Wei, G. Antichi, y B. Pfaff, ``Revisiting the Open vSwitch dataplane ten years later,'' en \emph{Proceedings of the ACM SIGCOMM 2021 Conference}, 2021.

\bibitem[12]{ref12}
``Configure MTU to maximize network performance,'' Tigera (Calico). [En línea]. Disponible en: \url{https://docs.tigera.io/calico/latest/networking/configuring/mtu}. [Consultado: 21-ago-2025].

\bibitem[13]{ref13}
``Service,'' Kubernetes. [En línea]. Disponible en: \url{https://kubernetes.io/docs/concepts/services-networking/service/}. [Consultado: 21-ago-2025].

\bibitem[14]{ref14}
Polito.it. [En línea]. Disponible en: \url{https://webthesis.biblio.polito.it/28635/1/tesi.pdf}. [Consultado: 21-ago-2025].

\bibitem[15]{ref15}
``Kubernetes networking: Comparative insights into API gateways and service mesh implementations,'' Aalto University, 2024.

\bibitem[16]{ref16}
M. Kotenko, D. Moskalyk, V. Kovach, y V. Osadchyi, ``Navigating the challenges and best practices in securing microservices architecture,'' \emph{CEUR-WS}, 2024.

\bibitem[17]{ref17}
``Roles and personas --- Kubernetes Gateway API,'' Kubernetes SIGs. [En línea]. Disponible en: \url{https://gateway-api.sigs.k8s.io/concepts/roles-and-personas/}. [Consultado: 21-ago-2025].

\bibitem[18]{ref18}
``CNI,'' \emph{cni.dev}. [En línea]. Disponible en: \url{https://www.cni.dev/docs/spec/}. [Consultado: 21-ago-2025].

\bibitem[19]{ref19}
IEEE, ``IEEE 802.1Q.'' [En línea]. Disponible en: \url{https://standards.ieee.org/ieee/802.1Q/6844/}. [Consultado: 21-ago-2025].

\bibitem[20]{ref20}
``host-local IP address management plugin,'' \emph{cni.dev}. [En línea]. Disponible en: \url{https://www.cni.dev/plugins/current/ipam/host-local/}. [Consultado: 21-ago-2025].

\bibitem[21]{ref21}
``Port-mapping plugin,'' \emph{cni.dev}. [En línea]. Disponible en: \url{https://www.cni.dev/plugins/current/meta/portmap/}. [Consultado: 21-ago-2025].

\bibitem[22]{ref22}
B. Pfaff \emph{et al.}, ``The design and implementation of Open vSwitch,'' \emph{USENIX NSDI}, 2015.

\bibitem[23]{ref23}
``Antrea --- Network Flow Visibility,'' Antrea.io. [En línea]. Disponible en: \url{https://antrea.io/docs/v1.0.0/docs/network-flow-visibility/}. [Consultado: 21-ago-2025].

\bibitem[24]{ref24}
S. Rose, O. Borchert, S. Mitchell, y S. Connelly, ``Zero Trust Architecture,'' NIST Special Publication 800-207, 2020.

\bibitem[25]{ref25}
O. Borchert, G. Howell, A. Kerman, S. Rose, y M. Souppaya, ``Implementing a Zero Trust Architecture: High-level document,'' NIST, Gaithersburg, MD, 2025.

\bibitem[26]{ref26}
R. Chandramouli y Z. Butcher, ``Building secure microservices-based applications using service-mesh architecture,'' NIST, Gaithersburg, MD, 2020.

\bibitem[27]{ref27}
J. Zhang, P. Chen, Z. He, H. Chen, y X. Li, ``Real-time intrusion detection and prevention with neural network in kernel using eBPF,'' en \emph{Proc. 54th Annual IEEE/IFIP Int. Conf. on Dependable Systems and Networks (DSN)}, 2024, pp. 416--428.

\bibitem[28]{ref28}
``CNI Essentials: Kubernetes Networking under the Hood,'' Tetrate. [En línea]. Disponible en: \url{https://tetrate.io/blog/kubernetes-networking}. [Consultado: 02-sep-2025].

\bibitem[29]{ref29}
D. E. Sierra Guerrero y H. A. Alba Castro, ``Optimización Integral de Redes Kubernetes: Comparativa de CNIs, Automatización de NetworkPolicies y Reducción de Sobredimensionamientos,'' anteproyecto de trabajo de grado, Universidad Distrital Francisco José de Caldas, Facultad Tecnológica, Ingeniería Telemática, Bogotá, Colombia, 2025.

\end{thebibliography}
